---
title: "FastReg"
author: "Matt Wheeler, Shail Choksi, Adam Burkholder, John House, Deepak Mav"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: united
vignette: >
  %\VignetteIndexEntry{FastReg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

In Biostatistics, data science, and informatics, regression using generalized linear models (McCullagh 2019) is often the primary workhorse for scientists seeking to understand the relationship between measured covariates and observed responses.  In the era of “big data,” one may be required to compute thousands or millions of regressions.  Computing this large number of regressions requires significant computational resources, and this fact is exacerbated with logistic regression, whose solution requires an iterative algorithm verses a single solve that is standard with linear regression for continuous data.  

Traditional software solutions to regression analyses available in the programming language R (R Core Team 2023) are a one-at-a-time approach, which analyzes each dataset as a unique analysis. Though such analyses occur in a fraction of a second, their focus on a single dataset at a time creates code inefficiencies resulting in duplicating effort when many regressions are needed. This creates noticeable bottlenecks when millions of regressions are required.  For example, if part of the covariate matrix is the same across regressions, computer operations that require the allocation of matrix memory or matrix multiplication will be needlessly duplicated.

FastReg tries to solve this by using the iteratively weighted least squares (IRLS) algorithm to perform logistic regression. It assumes that the data to be analyzed are available in text format, and that each response is saved in column format (a standard GWAS format). FastReg converts the data to the HDF5 format as it allows for random access from multiple threads, memory efficiencies and faster read times. Once converted, we use the iteratively reweighted least squares algorithm to perform logistic regression. The algorithm typically converges within a handful of iterations, and for speed, we cap the number of iterations (defaults to six). The output contains the summary with significance p-value, the maximum likelihood estimate, and the covariance estimates.

# Pre-requisites and Installation

## Install BLAS

-   **Recommended**: [Intel OneAPI](https://www.intel.com/content/www/us/en/developer/tools/oneapi/onemkl-download.html "Intel OneAPI")
-   [OpenBLAS](https://github.com/OpenMathLib/OpenBLAS "OpenBLAS github")

## Windows

-   Install [**Rtools for Windows**](https://cran.r-project.org/bin/windows/Rtools/ "Rtools for Windows")

## MacOS

-   Ensure that **Xcode Command Line Tools** are installed. Open **Terminal** and type: **`xcode-select -p`**.
    -   If you see the following: **/Library/Developer/CommandLineTools** command line tools are installed
    -   If not, check for system updates: **`System Settings > General > Software Update`**
-   Install the latest [**gfortran** **compiler**](https://cran.r-project.org/bin/macosx/tools/ "gfortran compiler")

## Installing FastReg

```
# install dependencies
install.packages(c("Rcpp", "RcppArmadillo", "stats", "data.table", "BiocManager", "devtools"))
Biocmanager::install("rhdf5")
library(devtools)
# install the package from github
install_github("SciomeLLC/FastReg")
library(FastReg)
```

# Dataset preparation

Let's look at an example GWAS dataset of 500 individuals x 5000 POIs stored as text files:

-   testdata.bin.pheno.txt - contains 500 individual ids and the binary response for each

    ```{r include=TRUE}
    # Binary phenotype data
    read.csv("../data/testdata.bin.pheno.txt", sep = "\t", nrows = 10)
    ```

-   testdata.covar.txt - contains the individual ids, and 3 covariates (age, sex and treatment)

    ```{r include=TRUE}
    # Covariate data
    read.csv("../data/testdata.covar.txt", sep = "\t", nrows = 10)
    ```

-   testdata.poi.txt - contains 500 individual ids and 5000 Predictors of Interest
<details>
  <summary>Click to see contents (large output)</summary>
    ```{r eval=TRUE, include=TRUE}
    # POI data
    read.csv("../data/testdata.poi.txt", sep = "\t", nrows = 2)
    ```
</details>

## Converting to HDF5
FastReg provides functionality to convert the POI text dataset into HDF5 dataset(s). 
Lets create a HDF5 dataset using the poi text data:

```{r include=TRUE, eval=FALSE}
library(FastReg)
FastRegImport("testdata", "./FastRegData", delimiter="\t")
```

This will create n (where n is optimized according to your machine configuration) HDF5 datasets in the "FastRegData" directory.

# Logistic Regression
With the text file converted to HDF5 format, we are ready to run logistic regression on the full dataset.

```{r include=TRUE, eval=FALSE}
FastReg(
  regression.type="logistic",
  pheno.file="data/testdata.bin.pheno.txt",
  pheno.rowname.cols="ind",
  phenotype = "bin.resp",
  covar.file="data/testdata.covar.txt",
  covariates=c("age", "sex", "treatment"),
  covariate.type = c("numeric", "categorical", "categorical"),
  covariate.standardize = c(TRUE, FALSE, FALSE),
  covariate.levels = c("", "F,M", "Placebo,Test"),
  covariate.ref.level = c("", "F", "Placebo"),
  POI.file.dir = "./FastRegData"
  POI.type = "dosage",
  POI.effect.type = "additive",
  output.dir = "FastRegResults"
)
```
The parameters for the covariates define the structure, data type, standardization, levels and reference levels. These parameters are all a list of the same size and need to be defined in the same order for each. The regression will try and spawn n processes and process the dataset in parallel. Two files will be created in the "FastRegResults" directory - one for the convergence and the other with the results of the logistic regression.

```{r include=TRUE, eval=FALSE}
read.csv("FastRegResults/Full_Convergence_stratum_1.tsv", delimiter="\t", nrows=10)
read.csv("FastRegResults/Full_Results_stratum_1.tsv", delimited="\t", nrwos=10)
```
